Incorporating Periodic Variability in Hidden Markov Models for Animal Movement

### 2016-09-21 bmc_template/manuscript.tex
putting figs at the appropriate place, going to replace sim results with a table

%% fig 1 goes somewhere here
The simulation experiment supports our hypothesis that homogeneous transition HMMs overestimate the number of hidden states when the model is misspecified (Figure~1). Heterogeneous transition models can always predict the correct number of states (in 100/100 simulations, BIC correctly identifies $n=2$ as the number of states), whereas the temporally homogeneous models overestimate the number of states (the correct value, $n=2$, is chosen most often, but in fewer than half of the simulations; values up to $n=5$ are frequently chosen).

\bmb{caption needs a little work}

%% fig 2 goes somewhere here

@@

%% fig 3 goes somewhere here

@@

The estimated emission parameter values
(mean and standard deviation of the step length in each state)
are similar, for both homogeneous and heterogeneous
models, across all cats (Figure 7 shows a subset of cats). 
In general, the states with longer mean step lengths are
similar between homogeneous and heterogeneous models. For cats 14 and 15, the states with
the longest or next-longest mean step lengths have similar means
and standard deviations; for cats 1 and 2, three long-step states
in the homogeneous HMM appear to divide two long-step states in 
the heterogeneous HMM. For short-step states, the heterogeneous HMM
tends to identify a high-variance state, while the homogeneous HMM
picks up states with very short step lengths (questionable in any
case because we have not taken any special efforts to account for
GPS error).

@@

%% In our panther
%% study, it is really hard to interpret multiple latent states
%% biologically.  We have no way of knowing what panthers are actually
%% thinking (it is certainly more complex than being in one of a small
%% number of discrete latent states); we don't know the ``true'' number
%% of latent states, nor are we able to observe them directly, although
%% incorporating additional direct observations of behaviour (if
%% available) can at least partially address this problem
%% \cite{fryxell_multiple_2008}.  Three distinct movement states seem
%% biologically interpretable for Florida panthers according to van de
%% Kerk et al. \cite{kerk2015hidden}: Short step length suggests resting
%% states, intermediate step length a foraging state, and long step
%% length a traveling state.

@@

Finally, if our goal is actually to estimate the number of discrete
behavioural modes underlying observed behaviour --- keeping in mind
that these discrete behavioural states are certainly an oversimplified
representation of animals' real internal states --- then, as we have
shown above, covariate-dependent models will generally be required to
avoid overestimating the number of states.  We note that researchers
in cluster analysis (of which HMMs are a special case) have shown that
the technical conditions required for BIC to apply may be violated
\cite{biernacki2000assessing}; however, BIC can be useful as an
approximate \emph{upper} limit on the number of states. Various
solutions to this problem have been proposed, including the
``integrated classification likelihood'' (ICL)
\cite{biernacki2000assessing,celeux2008selecting}, as well as a simpler ``knee point''
method \cite{zhao_knee_2008} that looks for the cluster size that
corresponds to the largest change in BIC rather than to the smallest
overall BIC.  Dean \emph{et al.} \cite{dean2012behavioural} took a similar approach, but
based on the log-likelihood curve rather than the BIC. However, HMMs are 
a special case even for the various solutions. 
% ICL uses the likelihood of the  most probable state sequence via the Viterbi algorithm to account for relative  state occupancies between hidden states as a penalizing factor \cite{celeux2008selecting}, but it does not have the same properties as the overall unconditional likelihood, such as,  monotonic increase in likelihood as the number of hidden states increase. Furthermore,  both log-likelihood and BIC based ``knee point'' methods look at the changes corresponding to a ``linear'' increase in number of states rather than the non-linear increase in parameters corresponding to the linear increase in states.

@@

\section*{Figures}
  \begin{figure}[h!]
   \includegraphics[width=5in]{figure/sim_results-1}
  \caption{\csentence{Simulation Results} BIC-optimal state frequency for 2-6 state HMMs with and without covariate transition on 100 two-state hidden Markov models with covariate transitioning simulations.}
      \end{figure}

\clearpage

\clearpage

\clearpage

\clearpage

\clearpage

\begin{figure}[h!]
   \includegraphics[width=5in]{figure/r_msdlist-1}
  \caption{\csentence{State Identification Plot} Mean and standard deviation of  step length by state for BIC-optimal HMMs (homogeneous and heterogeneous with sinusoidal transition).
      \end{figure}

>>>>

  \begin{figure}[h!]
   \includegraphics[width=5in]{figure/sim_results-1}
  \caption{\csentence{Frequency of choosing BIC-optimal state on simulated data.} 45 out of 100 sets of simulated data from a 2-state HMM with BIC-optimal state frequency for 2-6 state HMMs with and without covariate transition on 100 two-state hidden Markov models with covariate transitioning simulations.}
      \end{figure}

The simulation experiment supports our hypothesis that homogeneous transition HMMs can overestimate the number of hidden states when the model is misspecified (Figure~1). Heterogeneous transition models can always predict the correct number of states (in 100/100 simulations, BIC correctly identifies $n=2$ as the number of states), whereas the temporally homogeneous models overestimate the number of states (the correct value, $n=2$, is chosen most often, but in fewer than half of the simulations; values up to $n=5$ are frequently chosen).

@@

@@

\begin{figure}[h!]
   \includegraphics[width=5in]{figure/r_msdlist-1}
  \caption{\csentence{State Identification Plot} Mean and standard deviation of  step length by state for BIC-optimal HMMs (homogeneous and heterogeneous with sinusoidal transition).
  }
      \end{figure}

The estimated emission parameter values
(mean and standard deviation of the step length in each state)
are similar, for both homogeneous and heterogeneous
models, across all cats (Figure 6 shows a subset of cats). 
In general, the states with longer mean step lengths are
similar between homogeneous and heterogeneous models. For cats 14 and 15, the states with
the longest or next-longest mean step lengths have similar means
and standard deviations; for cats 1 and 2, three long-step states
in the homogeneous HMM appear to divide two long-step states in 
the heterogeneous HMM. For short-step states, the heterogeneous HMM
tends to identify a high-variance state, while the homogeneous HMM
picks up states with very short step lengths (questionable in any
case because we have not taken any special efforts to account for
GPS error).

@@

@@

Finally, if our goal is actually to estimate the number of discrete
behavioural modes underlying observed behaviour --- keeping in mind
that these discrete behavioural states are certainly an oversimplified
representation of animals' real internal states --- then, as we have
shown above, covariate-dependent models will generally be required to
avoid overestimating the number of states.  We note that researchers
in cluster analysis (of which HMMs are a special case) have shown that
the technical conditions required for BIC to apply may be violated
\cite{biernacki2000assessing}; however, BIC can be useful as an
approximate \emph{upper} limit on the number of states. Various
solutions to this problem have been proposed, including the
``integrated classification likelihood'' (ICL)
\cite{biernacki2000assessing,celeux2008selecting}, as well as a simpler ``knee point''
method \cite{zhao_knee_2008} that looks for the cluster size that
corresponds to the largest change in BIC rather than to the smallest
overall BIC.  Dean \emph{et al.} \cite{dean2012behavioural} took a similar approach, but
based on the log-likelihood curve rather than the BIC. However, HMMs are 
a special case even for the various solutions.

@@
